Mining the human microbiome for clinical insight

### 2018-11-24 abstract.tex
Write chap1 intro to human microbiome

Human-associated microbial communities are essential for health and have been implicated in many diseases.
DNA sequencing technology has enabled the detailed characterization of these communities, leading to a rapid expansion in studies investigating relationships between the human microbiome and disease.
However, identifying clinically-relevant associations from microbiome datasets is complicated by the high dimensional nature of the data and variability of communities across people.
In this thesis, I describe three projects which overcome various analysis challenges to identify clinically-relevant associations between the human microbiome and disease.
In the first project, I present an analysis of lung, stomach, and oropharyngeal microbiomes of pediatric patients with aerodigestive symptoms.
I leverage data collected from multiple sites per patient to identify  clinically-actionable alterations in aerodigestive community relationships in patients with swallowing dysfunction and to discover new characteristics of the human lung and stomach microbiomes.
In the second project, I perform a meta-analysis of case-control gut microbiome datasets across 28 studies and 10 diseases.
By standardizing processing and analysis methods across many datasets, I find consistent disease-specific and shared patterns of associations which can inform therapeutic development and clinical treatment approaches.
In the third project, I describe a framework for performing rational donor selection in fecal microbiota transplant clinical trials.
In this framework, microbial associations identified from clinical and basic research are used to inform the donor selection strategy, increasing the likelihood of a successful clinical trial.
These projects demonstrate a variety of approaches for mining human microbiome data to identify clinically-relevant associations and discover new fundamental properities of human-associated microbial communities.
Together, this work suggests multiple avenues forward for translating findings from microbiome data analyses into clinical impact.

>>>>

Human-associated microbial communities are essential for health and have been implicated in many diseases.
DNA sequencing technology has enabled the detailed characterization of these communities, leading to a rapid expansion in studies investigating relationships between the human microbiome and disease.
However, identifying clinically-relevant associations from microbiome datasets is complicated by the high dimensional nature of the data and variability of results across studies.
In this thesis, I describe three projects which overcome various analysis challenges to identify clinically-relevant associations between the human microbiome and disease.
In the first project, I present an analysis of lung, stomach, and oropharyngeal microbiomes of pediatric patients with aerodigestive symptoms.
I leverage data collected from multiple sites per patient to identify  clinically-actionable alterations in aerodigestive community relationships in patients with swallowing dysfunction and to discover new characteristics of the human lung and stomach microbiomes.
In the second project, I perform a meta-analysis of case-control gut microbiome datasets across 28 studies and 10 diseases.
By standardizing processing and analysis methods across many datasets, I find consistent disease-specific and shared patterns of associations which can inform therapeutic development and clinical treatment approaches.
In the third project, I describe a framework for performing rational donor selection in fecal microbiota transplant clinical trials.
In this framework, microbial associations identified from clinical and basic research are used to inform the donor selection strategy, increasing the likelihood of a successful clinical trial.
These projects demonstrate a variety of approaches for mining human microbiome data to identify clinically-relevant associations and discover new fundamental properties of human-associated microbial communities.
Together, this work suggests multiple avenues forward for translating findings from microbiome data analyses into clinical impact.

### 2018-11-24 chap1.tex
Write chap1 intro to human microbiome

\section{Human-associated microbial communities are essential for health and implicated in disease}

Intro paragraph about the human microbiome relevance in health and disease. Make sure this paragraph is site-agnostic.

Discuss big data aspect of microbiome research: sequencing provides opportunities and challenges. In this thesis, we present three projects which overcome these challenges to provide useful clinical insight from mining microbiome data.

\section{Multi-site sampling of the aerodigestive microbiome allows for the identification of clinical associations}

@@

In Chapter 2, we present an analysis that leverages simultaneous sampling of patients to identify clinically-relevant characteristics distinguishing patients with swallowing dysfunction from those with normal swallow.

@@

Suggest new avenues for approaches to treating aspiration.

@@

>>>>

\section{The human microbiome in health and disease}

The human microbiome, made up of the microbes that live in and on our bodies, is essential for health and has been implicated in many diseases.
Almost all human body sites are colonized by microbial communities, ranging from the gut, the largest human-associated microbial community, to the lungs, which were for many years considered sterile (citation needed).
In health, these communities fight off or prevent infections, interact with and regulate host immune responses and metabolism, and metabolize xenobiotics or other compounds which are indigestible by the host, among other functions.
Microbes can also cause or exacerbate disease, as well as impact how we metabolize the drugs we take.
Perturbations in the composition or function of these communities have been implicated in many diseases, including metabolic, autoimmune, neurological, and respiratory conditions.
<Maybe a sentence about therapies?>
Although humans known about the human microbiome for thousands of years, recent academic research into this field has expanded significantly (citation needed).
Emerging research has been driven in part by the increasing accessibility of DNA sequencing technology, which can provide a detailed view of the bacteria in these communities without the need for time-consuming and difficult culturing experiments.

Excitement in this field from scientists, engineers, clinicians, and entrepreneurs has also been driven by the non-invasiveness of the research, large range and accessibility of potential interventions, and the increasing potential for clinical impact.
Many human-associated communities can be sampled non-invasively, facilitating research that is performed with real human patient populations rather than model organisms.
Interventions which can change the composition of microbial communities already exist and are accessible to patients: lifestyle changes such as diet can have large effects on the microbiome, and commonly used drugs like antibiotics and probiotics are already in wide usage.
Thus, identifying ways to repurpose existing interventions for targeted outcomes is attractive and exciting because the potential to implement them quickly is high.
Finally, the potential clinical impact of microbiome-based interventions is vast and growing.
Fecal microbiota transplants (FMT) for recurrent Clostridium infection have already proven themselves as an effective treatment: over 80% of patients with recurrent C. diff are cured after FMT.
Furthermore, many of the diseases in which the microbiome may play a role are associated with industrialization and are increasing in incidence worldwide.

Despite the increasing accessibility of microbiome research and potential for clinical impact, identifying clinically-relevant associations in microbiome studies remains challenging.
Microbiome datasets are high-dimensional, containing hundreds to thousands of bacterial species but frequently only sampling tens to hundreds of patients.
Microbial communities are also highly variable across people, and identifying  individual bacterial biomarkers that can distinguish health and disease across many patients is difficult.
Finally, microbiome datasets provide a window into associations but not causal relationships.
Researchers must perform follow-up mechanistic studies or FMT clinical trials to confirm the clinical relevance of associations.
In this thesis, we present three projects which overcome various data analysis challenges to provide useful clinical insight from mining microbiome datasets.
The following chapters each present a unique analysis of microbiome data, each highlighting a different perspective on harnessing microbiome data to move beyond statistical significance and identify clinically meaningful associations.

\section{Leveraging multi-site sampling to overcome variability in the aerodigestive microbiome}

The aerodigestive tract is comprised of the respiratory and upper gastrointestinal tracts and includes the oropharynx, lungs, and stomach.
Unlike other human body sites, the aerodigestive tract is unique in that all sites are connected and exchange material through bidirectional flow.

@@

In Chapter 2, we present an analysis that leverages simultaneous sampling of patients to identify clinically-relevant microbiome characteristics distinguishing patients with swallowing dysfunction from those with normal swallow.

@@

Suggest new avenues for approaches to treating aspiration and for analyzin future lung microbiome studies.

@@

\section{More words}

Identifying modifications in these drugs

Interventions which can change the composition of microbial communities are also relatively accessible: diet and lifestyle are known to have large effects on the human microbiome, and existing drugs like antibiotics and probiotics are already in wide usage.

Many possible ways to intervene to improve microbiome: lifestyle changes, antibiotics, probiotics

There is also an exciting possibility of repurposing existing interventions like antibiotics and probiotics for greater or novel therapeutic effects.

There is also strong precedent for microbial-based therapies, and the

allowing for studies to iterate between real human patients and model organisms

with the motivating goal

Discuss big data aspect of microbiome research: sequencing provides opportunities and challenges.

### 2018-11-24 chap2.tex
Write chap1 intro to human microbiome

\chapter{Aerodigestive paper}

@@

\includepdf[pages=-, pagecommand={\thispagestyle{plain}}, scale=1.0]{texmex/ms}

\clearpage
\begin{figure}[ht]
\centering
\includegraphics{texmex/fig/fig_1}
\caption*{{\bf Figure 1. OTU dynamics and possible bottle effects in a paired-sample experiment.}
Four histograms are shown, one each for each sample in the experiment:
control-before (above, black), control-after (above, red), experimental-before
(below, black), experimental-after (below, red). Each histogram shows how many
OTUs (logarithmic $y$-axes) have what number of associated reads (logarithmic
$x$-axis). No bin is shown for OTUs with zero counts. The dynamics of two OTUs
are shown: black arrows point to the abundance bin for OTUs in the ``before''
sample and red arrows point their abundance bins in the ``after'' samples.}
\end{figure}

\clearpage
\begin{figure}[ht]
\centering
\includegraphics{texmex/fig/fig_2}
\caption*{{\bf Figure 2. TPL distribution fits OTU abundance distributions in multiple ecosystems.}
Probability-probability plots comparing the empirical cumulative distribution
function (horizontal axis) with the theoretical cumulative probability of a TPL
distribution fit to each data set (vertical axis, black solid line). The first
ten data points are marked with vertical dashes: the first dash (furthest lower
left) represents the fraction of OTUs that have 1 read, the second dash
represents the fraction of OTUs with 2 or fewer reads, and so forth. The dotted
black line indicates a perfect fit of the TPL to the empirical distribution ($y = x$).
The theoretical cumulative probability of a simple lognormal distribution
(red line) is shown to emphasize the quality of the TPL fit. The ecosystems are
ocean water from this study (top left), wastewater sludge from this study (top
right), human stool (bottom left; Human Microbiome Project [HMP] sample), and
human vagina (bottom right; HMP sample). 99\% \textit{de novo} OTUs are shown for all
samples.}
\end{figure}

\clearpage
\begin{figure}[ht]
\centering
\includegraphics{texmex/fig/fig_3}
\caption*{{\bf Figure 3. OTUs in replicate units have correlated dynamics.}
The dynamics of OTUs (circles) in three replicate bioreactors (replicate 1,
$x$-axis; replicates 2 and 3, $y$-axes) inoculated with the same material and
subjected to the same conditions. The dotted line ($y=x$) indicates a perfect
correlation: an OTU on this line would have exactly the same $\Delta z$ in both
replicates, while deviations show differences in dynamics. For example, in the
left plot, OTUs above the dotted line experienced a greater increase in
abundance in replicate 2 than in replicate 1 (or, a smaller decrease in 2 than
in 1), while OTUs below the line ``grew more'' in replicate 2 than in replicate 1
(or, ``died less'' in 2 than in 1). OTUs with infinite $\Delta z$ are plotted on the
plot's borders (e.g., the points in the lower-right corner of the first plot
represent OTUs that have $\Delta z = +\infty$ in replicate 1 and $\Delta z = -\infty$ in replicate 2).}
\end{figure}

\clearpage
\begin{figure}
\centering
\includegraphics{texmex/fig/fig_4}
\caption*{{\bf Figure 4. OTUs that respond to oil appear in five clades.}
On a phylogenetic tree built from the 16S sequences, organisms potentially
responding to crude oil are marked with open circles. OTUs that satisfy the $\Delta z$
criteria are marked with blue circles, OTUs that satisfy the $\Delta F$ criteria are
marked with orange circles, and OTUs that satisfy both are marked with black
circles. Information about the taxonomy and dynamics of these sequences are
shown in Table 1. The five clades (A through E) are labeled, and select
taxonomic groups are labeled to help orient the reader. The Archaea branch is
truncated. Scale bar: substitutions per site.}
\end{figure}

\clearpage
\begin{table}
\caption*{{\bf Table 1. OTUs with dynamic behavior in response to amendment with oil}
(\textit{on next page}).
All OTUs that satisfied the $\Delta z$ or $\Delta F$ criteria are listed. The first
three columns show taxonomy. The most specific RDP taxonomic
classification with at least 80\% bootstrap support is shown. The next
two columns indicate whether the OTU satisfied the $\Delta z$ criteria, the $\Delta F$
criteria, or both. The next six columns show the changes in relative
abundance ($\Delta \mathrm{r.a.}$), rescaled reads $z$, and cumulative distribution
function $F$ in the control (``ct'') and experimental (``ex'') units. The
value $\Delta z = \mathrm{n.a.}$ is shown for OTUs that had zero counts at both
timepoints in that microcosm; $\Delta z = \infty$ is shown for OTUs had zero counts
before the treatment and more than zero counts after the treatment.}
\end{table}

\clearpage
\begin{landscape}
\begin{table}
\centering
\begin{tabular}{*{12}{l}}
\toprule
 &  &  & \multicolumn{2}{c}{criteria} & \multicolumn{2}{c}{$\Delta\mathrm{r.a.}$} & \multicolumn{2}{c}{$\Delta z$} & \multicolumn{2}{c}{$\Delta F$} &  \\
\cmidrule(r){4-5} \cmidrule(r){6-7} \cmidrule(r){8-9} \cmidrule(r){10-11}
Clade & Classification & support & $\Delta z$ & $\Delta F$ & ct & ex & ct & ex & ct & ex & OTU ID \\
\midrule
A & \textit{Maricurvus} & 0.95 & * &  & 0.033 & 0.739 & 1.382 & 1.419 & $-0.0002$ & 0.0041 & 3 \\
 & \textit{Maricurvus} & 0.87 &  & * & 0 & 0.008 & n.d. & $\infty$ & 0 & 0.9958 & 63 \\
 & $\gamma$-Proteobacteria & 1 &  & * & 0 & 0.0033 & n.d. & $\infty$ & 0 & 0.9942 & 107 \\
 & \textit{Maricurvus} & 0.87 &  & * & 0 & 0.003 & n.d. & $\infty$ & 0 & 0.9939 & 111 \\
 & \textit{Maricurvus} & 0.96 & * & * & 0 & 0.0013 & 0.827 & $\infty$ & 0.1429 & 0.9883 & 119 \\
 & \textit{Maricurvus} & 0.97 & * &  & $-0.0002$ & 0.0002 & 0.115 & $\infty$ & 0.0263 & 0.9292 & 256 \\
 & \textit{Maricurvus} & 0.9 & * &  & $-0.0001$ & 0.0005 & 0.402 & $\infty$ & 0.1649 & 0.9659 & 262 \\
 & \textit{Maricurvus} & 0.94 &  & * & 0 & 0.0007 & n.d. & $\infty$ & 0 & 0.9767 & 291 \\
B & \textit{Pseudomonas} & 1 & * &  & 0.0027 & 0.0629 & 1.137 & 1.073 & 0.014 & 0.0063 & 14 \\
 & \textit{Pseudomonas} & 0.99 &  & * & 0 & 0.0142 & n.d. & $\infty$ & 0 & 0.9961 & 42 \\
 & \textit{Pseudomonas} & 0.91 &  & * & 0 & 0.0112 & n.d. & $\infty$ & 0 & 0.996 & 53 \\
 & Pseudomonaceae & 0.82 &  & * & 0 & 0.0045 & n.d. & $\infty$ & 0 & 0.995 & 88 \\
 & \textit{Pseudomonas} & 0.99 &  & * & 0 & 0.0025 & n.d. & $\infty$ & 0 & 0.9931 & 120 \\
 & \textit{Pseudomonas} & 0.91 &  & * & 0 & 0.0018 & n.d. & $\infty$ & 0 & 0.9913 & 167 \\
 & \textit{Pseudomonas} & 1 &  & * & 0 & 0.0016 & n.d. & $\infty$ & 0 & 0.9903 & 174 \\
C & \textit{Alcanivorax} & 1 & * & * & $-0.0001$ & 0.0004 & 0.402 & $\infty$ & 0.1649 & 0.9599 & 206 \\
 & \textit{Alcanivorax} & 1 & * &  & $-0.0007$ & $-0.0001$ & 0.007 & $-0.143$ & $-0.0143$ & 0.0148 & 270 \\
D & \textit{Methylophaga} & 1 &  & * & 0 & 0.0006 & n.d. & $\infty$ & 0 & 0.9739 & 210 \\
E & Rhodobacteraceae & 1 & * & * & $-0.0003$ & 0.0008 & $-0.167$ & $\infty$ & $-0.1188$ & 0.9799 & 104 \\
 & Rhodobacteraceae & 1 & * & * & 0.0003 & 0.0003 & 1.237 & $\infty$ & 0.2935 & 0.9384 & 105 \\
 & Rhodobacteraceae & 1 & * & * & $-0.0002$ & 0.0001 & 0.115 & $\infty$ & 0.0263 & 0.7678 & 226 \\
 & Rhodobacteraceae & 1 &  & * & 0 & 0 & n.d. & $\infty$ & 0 & 0.6098 & 288 \\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

\clearpage
\begin{figure}[ht]
\centering
\includegraphics{texmex/fig/fig_s1}
\caption*{{\bf S1 Fig. TPL fits OTUs called by different methods.}
Probability-probability plots
comparing the empirical cumulative distribution function ($x$-axis) with the
theoretical cumulative probability of a TPL distribution fit to the
distribution of OTUs computed using different OTU-calling methods ($y$-axis,
black line). This is same ocean sample as in Figures 1 and 2. The first ten
data points are marked with vertical dashes: the first dash (furthest lower
left) represents fraction of OTUs with 1 read, the second dash represents the
fraction of OTUs with 2 or fewer reads, and so forth. The dotted black line
indicates a perfect fit ($y = x$). The theoretical cumulative probability of a
simple lognormal distribution fit to each OTU distribution (red) is shown to
emphasize the quality of the TPL fit. The methods are unique sequences (i.e.,
100\% identity OTUs; top left), 97\% reference-based OTUs from Greengenes (top
right), \textit{de novo} 97\% OTUs (bottom left), and genus-level OTUs computed with RDP
(bottom right). The empirical goodness-of-fit test described in the main text
yields $p = 0.35, 0.40, 0.44, 0.41$ for these data.}
\end{figure}

\clearpage
\begin{figure}[ht]
\centering
\includegraphics{texmex/fig/fig_s2}
\caption*{{\bf S2 Fig. OTU dynamics measured by $\Delta z$.}
Each OTU present in the microcosm experiment described in the main text is
shown. OTUs that meet the $\Delta z$ criterion described in the text ($\Delta z$ in treatment >
$\Delta z$ in control $- 0.5$) are in red. OTUs that meet the criterion and are among
the 304 most abundant sequences in the four microcosm experiments (i.e., those
shown as blue dots in Figure 4) are circled. OTUs with an undefined $\Delta z$ value in
either microcosm are not shown, while OTUs with infinite $\Delta z$ values are shown at
the border of the figure (e.g., an OTU with $\Delta z = +\infty$ in the control unit and $-\infty$
in the experimental unit would be shown in the lower-left corner).}
\end{figure}

\clearpage
\begin{figure}[ht]
\centering
\includegraphics{texmex/fig/fig_s3}
\caption*{{\bf S3 Fig. OTU dynamics measured by $\Delta F$.}
Each OTU present in the microcosm experiment described in the main text is
shown. OTUs that meet the $\Delta F$ criteria described in the text ($\Delta F$ in treatment >
0.5; $\Delta F$ in control $< 0.5$) are in red. OTUs that meet those criteria and are
among the 304 most abundant sequences in the four microcosm experiments (i.e.,
those shown as red dots in Figure 4) are circled.}
\end{figure}

>>>>

\chapter{Multi-site aerodigestive sampling reveals altered microbial exchange between lung, oropharyngeal, and gastric microbiomes in children with impaired swallow function}

@@

Claire Duvallet, Kara Larson, Scott Snapper, Sonia Iosim, Ann Lee, Katherine Freer, Kara May, Eric J. Alm, and Rachel Rosen. "Aerodigestive sampling reveals altered microbial exchange between lung, oropharyngeal, and gastric microbiomes in children with impaired swallow function." In review, November 2018.

### 2018-11-24 cover.tex
Write chap1 intro to human microbiome

\title{Mining the human microbiome for clinical insight and applications}

@@

Mentors:
Rafa - will always be a friend of the lab
Jim
Deb

Thomas - being my first introduction to computational work with patience and enthusiasm, enabling me to find my passion here
Mariana - drive and ability to make impossible things actually happen was an inspiration, expanded the scope of what a PhD student could do
Scott - being a role model, mentor, collaborator, and now close friend. Hope you remain all four forever. Also for encouraging me to start doing diversity and REFS
Ilana - for being the mentor I didn't know I needed
Sean Gibbons - modeling pure interest in science for the sake of science, encouraging me and being patient when I thought it was all fake, modeling success and being a joy to work with

>>>>

\title{Mining the human microbiome for clinical insight}

@@

Mentors:
Rafa - will always be a friend of the lab
Jim
Deb
Ilana - for being the mentor I didn't know I needed

Thomas - being my first introduction to computational work with patience and enthusiasm, enabling me to find my passion here
Mariana - drive and ability to make impossible things actually happen was an inspiration, expanded the scope of what a PhD student could do
Scott - being a role model, mentor, collaborator, and now close friend. Hope you remain all four forever. Also for encouraging me to start doing diversity and REFS
Sean Gibbons - modeling pure interest in science for the sake of science, encouraging me and being patient when I thought it was all fake, modeling success and being a joy to work with
